Liquid Amortization: Proving Amortized Complexity with LiquidHaskell (Functional Pearl)

### 2024-07-18 paper/main.tex
Add more review comments

Usually formal proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses. This translation step has a few disadvantages. First, the semantics of the target language could differ from the semantics of the production language. This is especially true for Haskell which is lazy, while most theorem provers have a strict language semantics. Second, such a translation step could introduce subtle differences or bugs, either by accident when doing a manual translation or through a bug when an automated translation tool is used. Lastly, the target language may not support certain features present in the source language (e.g. polymorphic recursion in data types, see section~\ref{sec:fingertree}). In those cases, the missing features have to be emulated in the target language, again opening the door for small and subtle errors. Because the code that is actually used in the formal proofs differs from the source code that is running in production, it is often hard to reconnect the proofs back to the original production code.

Another possibility is using LiquidHaskell~\cite{liquidhaskell}. It is a refinement type~\cite{refinement_types} checker plugin\footnote{Before 2020, LiquidHaskell was implemented as a standalone executable; it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in industry. It allows specifying additional constraints on types, e.g. "integers greater than zero" instead of just "integers". These constraints are propagated through the code and checked by an external SMT solver. With a handful of proof combinators defined by LiquidHaskell, it is possible to use Haskell directly as a theorem prover, enabling intuitive equational reasoning within the language~\cite{tpfa}. This avoids any translation steps and makes it easy to keep implementation and its respective proof in sync, as any failure to do so will result in a compiler error. Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the same as the proof language.

@@

There are multiple techniques to prove amortized complexity; we present the two most popular ones. Note that both of these techniques only apply in a non-shared setting where intermediate results are not reused. If we were allowed to keep a reference to the original data structure after an operation, we could do the same expensive operation several times without "paying" for it. Some data structures can exploit laziness to have the same (good) amortized complexity also in a shared setting, however formally verifying this would require explicit reasoning about the lazy semantics of the language and thus massively complicate the analysis.

@@

\begin{lstlisting}
multipopP k xs@(Elem x s) = multipopT k xs
    + phi (multipop k xs) - phi xs
-- Use definition of multipopT for recursive case
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop k xs)) - phi xs
-- Use definition of multipop for recursive case
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop (k - 1) s)) - phi xs
-- Use definition of phi (Elem _ _)
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop (k - 1) s)) - (1 + phi s)
-- Remove parentheses
  === 1 + multipopT (k - 1) s
    + phi (snd (multipop (k - 1) s)) - 1 - phi s
-- 1s cancel out
  === multipopT (k - 1) s
    + phi (snd (multipop (k - 1) s)) - phi s
-- use induction hypothesis
  ? multipopP (k - 1) s
  *** QED
\end{lstlisting}

@@

{-@ automatic-instances insertTreeP @-}
{-@ insertTreeP :: t:Tree a -> f:Heap a ->
  { insertT t f + pot (insertTree t f)
    - pot f <= 2 }
@-}
insertTreeP :: Ord a => Tree a -> Heap a -> Proof
insertTreeP t (F1 t' f') = trivial
  ? insertTreeP (mergeTree t t') f'
insertTreeP _ _ = trivial
\end{lstlisting}

@@

Now for the actual proof we will only show one of the base cases as the others follow the same schema. All the base cases have a constant upper bound on their complexity. For the recursive case we also need some small helper facts about the logarithm such as monotonicity.

@@

{-@ glueAmortized :: q1:Seq a
  -> { as:[a] | len as <= 3 }
  -> q2:Seq a -> {
      glueT q1 as q2 + pot (glue q1 as q2)
      - pot q1 - pot q2
    <= log2 (max (len (seqToList q1)
      + len (seqToList q2)) 2) + 14
  } @-}
glueAmortized :: Seq a -> [a] -> Seq a -> Proof
glueAmortized q1@Nil as q2 =
  glueT q1 as q2 + pot (glue q1 as q2)
    - pot q1 - pot q2
  === 1 + foldrT cons consT q2 as
    + pot (foldr cons q2 as) - pot q1 - pot q2
  ? foldrTCons q2 as
  =<= 1 + 3 * length as + 1 - pot q1
  =<= log2 (max (length (seqToList q1)
    + length (seqToList q2)) 2) + 14
  *** QED
glueAmortized qq1@(More u1 q1 v1) as
  qq2@(More u2 q2 v2) =
  glueT qq1 as qq2 + pot (glue qq1 as qq2)
  - pot qq1 - pot qq2
  === 1 + glueT q1 (toTuples
        (toList v1 ++ as ++ toList u2)) q2
      + pot (More u1 (glue q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2) v2)
      - danger u1 - pot q1 - danger v1 - danger u2
      - pot q2 - danger v2
  === 1 + glueT q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2
      + pot (glue q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2)
      - pot q1 - danger v1 - danger u2 - pot q2
-- apply induction
  ? glueAmortized q1
      (toTuples (toList v1 ++ as ++ toList u2)) q2
  =<= log2 (max (length (seqToList q1)
        + length (seqToList q2)) 2) + 15
-- move (+1) into the logarithm
  ? divCancel (max (length (seqToList q1)
      + length (seqToList q2)) 2)
  === 1 + log2 (2 * max (length (seqToList q1)
      + length (seqToList q2)) 2 `div` 2) + 14
  === log2 (2 * max (length (seqToList q1)
      + length (seqToList q2)) 2) + 14
-- simplify
  ? log2Mono (2 * max (length (seqToList q1)
      + length (seqToList q2)) 2)
    (4 + 2 * length (seqToList q1)
      + 2 * length (seqToList q2))
  =<= log2 (4 + 2 * length (seqToList q1)
      + 2 * length (seqToList q2)) + 14
  ? log2Mono (4 + 2 * length (seqToList q1)
    + 2 * length (seqToList q2))
      (4 + length (tuplesToList (seqToList q1))
        + length (tuplesToList (seqToList q2)))
  =<= log2 (4 + length (tuplesToList (seqToList q1))
      + length (tuplesToList (seqToList q2))) + 14
  ? log2Mono (4 + length (tuplesToList (seqToList q1))
    + length (tuplesToList (seqToList q2))) 
      (max (length (toList u1 ++ tuplesToList
          (seqToList q1) ++ toList v1)
        + length (toList u2 ++ tuplesToList
            (seqToList q2) ++ toList v2)) 2)
  =<= log2 (max (length (toList u1 ++ tuplesToList
        (seqToList q1) ++ toList v1)
      + length (toList u2 ++ tuplesToList
        (seqToList q2) ++ toList v2)) 2) + 14
  === log2 (max (length (seqToList qq1)
      + length (seqToList qq2)) 2) + 14
  *** QED
\end{lstlisting}

>>>>

Usually formal proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses. Such a translation step could introduce subtle differences or bugs if this translation is done manually. Additionally, the target language may not support certain features present in the source language (e.g. polymorphic recursion in data types, see section~\ref{sec:fingertree}). In those cases, the missing features have to be emulated in the target language, again opening the door for small and subtle errors. Because the code that is actually used in the formal proofs differs from the source code that is running in production, it is often hard to reconnect the proofs back to the original production code.

Another possibility is using LiquidHaskell~\cite{liquidhaskell}. It is a refinement type~\cite{refinement_types} checker plugin\footnote{Before 2020, LiquidHaskell was implemented as a standalone executable; it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in industry. It allows specifying additional constraints on types, e.g. "integers greater than zero" instead of just "integers". These constraints are propagated through the code and checked by an external SMT solver. With a handful of proof combinators defined by LiquidHaskell, it is possible to use Haskell directly as a theorem prover, enabling intuitive equational reasoning within the language~\cite{tpfa}. This makes it easy to keep the implementation and its respective proof in sync, as any failure to do so will result in a compiler error. LiquidHaskell errors are also integrated with the normal Haskell IDE tooling, so the user can see proof failures directly in their editing environment. Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the same as the proof language.

@@

There are multiple techniques to prove amortized complexity; we present the two most popular ones. Note that both of these techniques only apply in a non-shared setting where intermediate results are not reused. If we were allowed to keep a reference to the original data structure after an operation, we could do the same expensive operation several times without "paying" for it. Some data structures can exploit laziness to have the same (good) amortized complexity also in a shared setting, however formally verifying this would require explicit reasoning about the lazy semantics of the language and thus massively complicate the analysis. Note that in a non-shared setting the amortized complexity bounds apply regardless of whether the language is lazy or strict. Laziness may cause the data structure to do less work -- and thus have a better complexity -- but as an upper bound the amortized complexity is still valid.

@@

\begin{lstlisting}
multipopP k xs@(Elem x s) = multipopT k xs
    + phi (multipop k xs) - phi xs
-- Use definition of multipopT n (Elem _ s)
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop k xs)) - phi xs
-- Use definition of multipop n (Elem _ s)
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop (k - 1) s)) - phi xs
-- Use definition of phi (Elem _ s)
  === (1 + multipopT (k - 1) s)
    + phi (snd (multipop (k - 1) s))
    - (1 + phi s)
-- Remove parentheses
  === 1 + multipopT (k - 1) s
    + phi (snd (multipop (k - 1) s)) - 1 - phi s
-- 1s cancel out
  === multipopT (k - 1) s
    + phi (snd (multipop (k - 1) s)) - phi s
-- use induction hypothesis
  ? multipopP (k - 1) s
  *** QED
\end{lstlisting}

@@

{-@ automatic-instances insertTreeP @-}
{-@ insertTreeP :: t:Tree a -> f:Heap a ->
  { insertT t f + phi (insertTree t f)
    - phi f <= 2 }
@-}
insertTreeP :: Ord a => Tree a -> Heap a -> Proof
insertTreeP t (F1 t' f') = trivial
  ? insertTreeP (mergeTree t t') f'
insertTreeP _ _ = trivial
\end{lstlisting}

@@

Now for the actual proof we will only show one of the base cases as the others follow the same schema. All the base cases have a constant upper bound on their complexity. For the recursive case we also need some small helper facts about the logarithm such as monotonicity. We also use a \texttt{where} binding to factor out common code and make the proof more readable.

@@

{-@ glueAmortized :: q1:Seq a
  -> { as:[a] | len as <= 3 }
  -> q2:Seq a -> {
      glueT q1 as q2 + pot (glue q1 as q2)
      - pot q1 - pot q2
    <= log2 (max (len (seqToList q1)
      + len (seqToList q2)) 2) + 14
  } @-}
glueAmortized :: Seq a -> [a] -> Seq a -> Proof
glueAmortized q1@Nil as q2 =
  glueT q1 as q2 + pot (glue q1 as q2)
    - pot q1 - pot q2
  === 1 + foldrT cons consT q2 as
    + pot (foldr cons q2 as) - pot q1 - pot q2
  ? foldrTCons q2 as
  =<= 1 + 3 * length as + 1 - pot q1
  =<= log2 (max (length (seqToList q1)
    + length (seqToList q2)) 2) + 14
  *** QED
glueAmortized qq1@(More u1 q1 v1) as
  qq2@(More u2 q2 v2) =
  glueT qq1 as qq2 + pot (glue qq1 as qq2)
  - pot qq1 - pot qq2
  === 1 + glueT q1 (toTuples
        (toList v1 ++ as ++ toList u2)) q2
      + pot (More u1 (glue q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2) v2)
      - danger u1 - pot q1 - danger v1 - danger u2
      - pot q2 - danger v2
  === 1 + glueT q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2
      + pot (glue q1 (toTuples
          (toList v1 ++ as ++ toList u2)) q2)
      - pot q1 - danger v1 - danger u2 - pot q2
-- apply induction
  ? glueAmortized q1
    (toTuples (toList v1 ++ as ++ toList u2)) q2
  =<= log2 (max n 2) + 15
-- move (+1) into the logarithm
  ? divCancel (max n 2)
  === 1 + log2 (2 * max n 2 `div` 2) + 14
  === log2 (2 * max n 2) + 14

-- simplify using monotonicity of log2
  ? log2Mono (2 * max n 2) (4 + 2 * n1 + 2 * n2)
  =<= log2 (4 + 2 * n1 + 2 * n2) + 14
  ? log2Mono (4 + 2 * n1 + 2 * n2) (4 + n1' + n2')
  =<= log2 (4 + n1' + n2') + 14
  ? log2Mono (4 + n1' + n2')
    (max (length (toList u1
      ++ tuplesToList (seqToList q1) ++ toList v1
    ) + length (toList u2
      ++ tuplesToList (seqToList q2) ++ toList v2
    )) 2)
  =<= log2 (max (length (toList u1
      ++ tuplesToList (seqToList q1) ++ toList v1)
    + length (toList u2
      ++ tuplesToList (seqToList q2) ++ toList v2)
    ) 2) + 14
  === log2 (max (length (seqToList qq1)
    + length (seqToList qq2)) 2) + 14
  *** QED
  where
    n1 = length (seqToList q1)
    n2 = length (seqToList q2)
    n = n1 + n2
    n1' = length (tuplesToList (seqToList q1))
    n2' = length (tuplesToList (seqToList q2))
\end{lstlisting}
